---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(knitr)
library(tinytex)
```

## Loading and preprocessing the data
The data is ingested by first downloading the .zip file from the url provided in the assignment. The first time the code is run the file is downloaded and unzipped in to separate folder "/activity_data". If the download and extraction has already been done, the file is not downloaded again.
```{r 'data file download'}
dataset_url <- paste0("https://d396qusza40orc.cloudfront.net/",
                      "repdata%2Fdata%2Factivity.zip")
downloaded_zipfile <- "./activity.zip"

if (!file.exists(downloaded_zipfile)) {
        download.file(dataset_url, destfile = downloaded_zipfile, 
                      method = "curl", mode = "wb")
}
unzipped_file_folder <- "./activity_data"

if (!file.exists(unzipped_file_folder)) {
        unzip(downloaded_zipfile, exdir = unzipped_file_folder,
              unzip = "internal")
}
```
With the data file successfully downloaded, the 'activity.csv' file is read into the 'activity' variable.
```{r 'reading data into r' }
activity <- read.csv(file = paste0(unzipped_file_folder, "/", "activity.csv"))
```
One of the first checks performed was to access the prevalence of 'NAs' in the step data.
```{r 'calculate overall fraction of and total numberof NAs', echo=FALSE}
(activity %>% summarize(overall_frac_NA = mean(is.na(steps)),
                       total_NAs = sum(is.na(steps))))
```
The 'activity' dataframe contains a total of 2304 NAs that accounts for ~13% of all of the rows of step data. We can look a little closer at the impact of these NAs by checking how they are distributed across the 5-minute time intervals.

```{r}
(NA_by_interval <- activity %>%
        group_by(interval) %>%
        summarize(frac_NA = mean(is.na(steps)), count_NA = sum(is.na(steps))))
```
From the first ten rows of intervals we can see that each interval has eight NAs. With 288 5-minute intervals and 8 NAs per intervals that gives the `r 288 * 8` total number of NAs we found above. It looks like the NAs are uniformly spread across the intervals and do not need any additional consideration.

## What is mean total number of steps taken per day?
The mean total number of daily steps is calculated by grouping the activity dataframe by the 'date' variable and summarizing the sum of the daily steps with the NA values removed from the sum.
```{r 'total steps per day w/o NAs'}
total_steps_per_day <- activity %>%
        group_by(date) %>%
        summarize(daily_step_total=sum(steps, na.rm = TRUE))
```
With the 'total_steps_per_day' (this excludes NA values), we can create a histogram to get an idea of how the number of total daily steps varies. Note: I tried a few different bin widths and decided that 15 bins did an adequate job of showing the overall pattern of variation.
```{r 'histogram of total steps per day w/o NAs'}
hist(total_steps_per_day$daily_step_total,
     breaks = 15,
     main = "Total Steps per Day",
     xlab = "Total Steps per Day (NAs removed)")
```

Two important statistics for this data are the mean and median daily step count.
```{r 'mean and median steps per day w/o NAs'}
(daily_step_stats <-total_steps_per_day %>%
        summarize(mean_daily_steps = mean(daily_step_total),
                  median_daily_steps = median(daily_step_total)))
```
The mean (`r format(daily_step_stats$mean_daily_steps, digits = 0)` steps) is less than the median (`r format(daily_step_stats$median_daily_steps, digits = 0)` steps), but not so much as to raise concerns about the step data being significantly skewed.

## What is the average daily activity pattern?
We now want to turn our attention to how the step count varies across the 288 5-minute intervals in each day. This can be accomplished by grouping the activity dataframe by 'interval' and then creating summary values of the total and average steps per interval across all 61 days in the dataframe. In this analysis, the NA values are excluded.
```{r 'calculate total and avg steps per interval w/o NAs'}
avg_steps_by_interval <- activity %>%
        group_by(interval) %>%
        summarize(tot_steps = sum(steps, na.rm=TRUE),
                  avg_steps = mean(steps, na.rm=TRUE))
```
The overall pattern of average steps per 5-minute interval can be seen in the following time series plot.
```{r 'time series plot of avg steps per interval w/o NAs'}
with(avg_steps_by_interval, plot(interval, avg_steps,
        type = "l",
        main = "Average Steps per 5-minute Interval",
        sub = "(NAs removed)",
        xlab = "5-min Interval",
        ylab = "Average Step Count"))
```

The maximum average step count in an interval is calculated by finding the row with the highest average step count. (again noting that we have excluded the NAs in this view of the data)
```{r 'find the interval with the max avg step count w/o NAs'}
max_avg_steps_by_interval <- activity %>%
        group_by(interval) %>%
        summarize(avg_steps = mean(steps, na.rm=TRUE)) %>%
        slice_max(order_by = avg_steps)
```
The highest average step count (`r format(max_avg_steps_by_interval$avg_steps, digits=1)` steps) is found in the interval `r max_avg_steps_by_interval$interval`. It looks like a lot of the daily step total occurs in the morning wthe peak around 8:35am.

## Imputing missing values
Since ~13% of the step count data is missing, we can consider replacing the NA values with a meaningful substitute value. We start by creating a new dataframe to include imputed step counts.
```{r 'create new dataframe for imputed steps'}
imputed_activity <- activity %>% add_column(imputed_steps = NA)
```
For this analysis, I chose to replace the missing values with the average step counts calculated for each 5-minute interval (with the NAs removed). This seems reasonable since the average step count for each interval was quite variable. If the original step count for an interval was missing, the average step count for that interval replaced the NA in the 'imputed_steps' variable. If there was a step count value in the original 'steps' variable it was used in the 'imputed_steps' variable.
```{r 'create imputed step counts for each interval'}
for (i in 1:nrow(imputed_activity)) {
        if (is.na(imputed_activity$steps[i])) {
                imputed_activity$imputed_steps[i] <-
                        avg_steps_by_interval$avg_steps[activity$interval[i] ==
                                                avg_steps_by_interval$interval]
        }
        else {
                imputed_activity$imputed_steps[i] <- imputed_activity$steps[i]
        }
}

rm(i)
```
With the imputed steps we can now calculate the total daily step count by summarizing the new data by 'date'.
```{r 'calculate total daily step count using imputed step counts'}
total_steps_per_day_imputed <- imputed_activity %>%
        group_by(date) %>%
        summarize(daily_step_total_imputed=sum(imputed_steps))
```
As shown in the following histogram, using the imputed step counts, the frequency in the first bin has gone down since there are fewer zeros in the imputed step data.
```{r 'histogram of daily step total using imputed step data'}
hist(total_steps_per_day_imputed$daily_step_total_imputed,breaks = 15,
     main = "Total Steps per Day",
     xlab = "Total Steps per Day (NAs replaced with interval mean)")
```

We can now also calculate the mean and median daily step counts using the imputed step data.
```{r calculate mean and median daily step count using imputed step counts}
daily_step_stats_imputed <-total_steps_per_day_imputed %>%
        summarize(mean_daily_steps_imputed = mean(daily_step_total_imputed),
                  median_daily_steps_imputed = median(daily_step_total_imputed))
```
The mean (`r format(daily_step_stats_imputed$mean_daily_steps_imputed, digits = 5)` steps) is now equal to the median (`r format(daily_step_stats_imputed$median_daily_steps_imputed, digits = 5)` steps). It would be interesting to see if this a genaral result when the average value is used to replace an NA. The mean (median) value for the imputed data set is larger than the result with the NAs excluded. This is not suprising since we replaced the NAs with the averages. 

## Are there differences in activity patterns between weekdays and weekends?
The final step in this analysis is to assess whether there are differences in the step-related activity patterns between weekdays (Monday to Friday) and weekends (Saturday and Sunday); using the imputed step data. This involves first creating a 'week_day' variable that relates the date to the day of the week. Since there are more weekdays than weekend days, I initialized the 'type_of_day' variable with the value 'Weekday'. Where the day of the week was a weekend day, the 'type_of_day' variable was updated to 'Weekend'.

```{r 'creating and populating the type_of_day variable to have levels of "Weekday" and "Weekend"'}
imputed_activity <- imputed_activity %>%
        mutate(week_day = weekdays(as.Date(date, "%Y-%m-%d")))

imputed_activity <- imputed_activity %>% add_column(type_of_day = "Weekday")

for (i in 1:nrow(imputed_activity)) {
        if (imputed_activity$week_day[i] %in% c("Saturday", "Sunday")) {
                imputed_activity$type_of_day[i] <- "Weekend"
        }
}

rm(i)
```
Since we want to compare the step-related activity patterns between weekdays and weekends, we can calculate the average steps per 5-minute interval for the weekday subset of the 'imputed-activity' and the weekend subset of the 'imputed_activity' dataframe. These two subsets are recombined (row-wise) to provide a single dataframe for creating time series plot dor each type of day.
```{r 'calculating the activity patterns for weekdays and weekends separately and then combining them'}
imputed_avg_steps_by_interval_weekday <- imputed_activity %>%
        filter(type_of_day=="Weekday") %>%
        group_by(interval) %>%
        summarize(tot_steps = sum(imputed_steps),
                  avg_steps = mean(imputed_steps)) %>%
        mutate(type_of_day = "Weekday")

imputed_avg_steps_by_interval_weekend <- imputed_activity %>%
        filter(type_of_day=="Weekend") %>%
        group_by(interval) %>%
        summarize(tot_steps = sum(imputed_steps),
                  avg_steps = mean(imputed_steps)) %>%
        mutate(type_of_day = "Weekend")

imputed_avg_steps_by_interval <- rbind(imputed_avg_steps_by_interval_weekday,
                                       imputed_avg_steps_by_interval_weekend)
```
A graphical comparison of the step-related activity patterns between weekdays and weekends is shown in the following figure with one panel for the 'Weekday' data and another for the 'Weekend' data.
```{r 'two panel time series plot of avg interval step data for Weekdays and Weekends'}
ggplot(imputed_avg_steps_by_interval, aes(interval, avg_steps)) +
        geom_line() + facet_grid(type_of_day ~ .) +
        ggtitle("Average Steps per 5-minute Interval") +
        theme(plot.title = element_text(hjust = 0.5)) +
        labs(x = "5-minute Interval",
             y = "Average Step Count",
             caption = "(NAs replaced with interval mean)")
```

One quick observation is that the spike in activity between 8:00 and 10:00am on weekdays is not as prominent in the weekend data and there seem to be more high step count intervals overall.